Multi-level modeling with Openflexo/\FML - A contribution to the \mpc

### 2021-05-31 8-discussion.tex
sync Overleaf -> Github
antidote discussion

  %plan plus ou moins imposé par l'appel
%\todo[inline]{mandatory discussion aspects ↓}

@@

  Our solution uses the basic modeling constructs depicted by the \FML core
  metamodel (Figure~\ref{fig:mm}) and described in
  section~\ref{sec:technology}. Models are \emph{virtual models}. They are
  composed of \emph{concepts}, being themselves concepts. Concepts may have
  \emph{roles} and \emph{behaviors} (actions one can perform).

  In order to provide the graphical tools, our solution also uses \emph{model
  slots} to connect some concept instances to external elements (in our case:
  instance of diagram from our diagramming TA). The use of slots has been
  described in section~\ref{sec:model} and is illustrated by bold circles in
  the Figure~\ref{fig:ToolingArchitecture}).

@@

  Thanks to the level-agnosticity nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation.

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models.  Note that in our approach, the
  metamodels are built in an ad hoc way together with the models
  (co-construction), therefore the user also validates on a ad hoc basis.

@@

  %\noteJC{pas trop clair pour moi pour le moment -> N/A ? }
  Due to the nature of our approach and its level-agnosticity, \emph{deep characterization} does not 
  apply to our solution. Such a mechanism could probably be encoded by specific behaviors, however 
  it would not be a generic mechanism, making it difficult to reuse for another problem.

@@

  Due to the level-agnostic aspect of our approach, we could insert a new
  level, for example. It would consist in creating new concepts we would link
  to other ones. Then, we would add any necessary constraints on those concepts
  and on the relationships linking them to the other existing concepts. This type 
  of change can be easily done. On the other hand, making changes to the existing 
  models could be difficult. For example, given a requirement, an instance cannot 
  be changed into a concept easily. This case could occur if the challenge had 
  refined the specification by requiring specific design tasks. In this context, 
  we would have to transform \emph{Design Task} into a concept whose instances 
  could have been "using agile methodology".

@@

%\todo[inline]{recommended discussion aspects ↓}
\subsection{Tools}

@@

As described in section~\ref{subsec:tooling}, our solution is fully tooled with
Openflexo.  We use the Openflexo infrastructure to build models and metamodels
together, and to provide dedicated tooling to edit a Process type, to enact a
process and to execute it. Being able to provide a solution and its associated
tools quickly and easily is a noticeable feature.  We could also have taken
advantage of the model federation to connect our solution to other tools
dedicated to process edition and to process execution (\eg a \BPMN engine).
However we already had all the necessary components to provide the
aforementioned graphical tools.

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realised by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked. Thus, a part of
the verification relies on the fact that users add behaviors when they build
the metamodels and the models.

>>>>

  %plan plus ou moins imposé par l'appel
%\todo[inline]{mandatory discussion aspects}

@@

  Our solution uses the basic modeling constructs depicted by the \FML core
  metamodel (Figure~\ref{fig:mm}) and described in
  the section~\ref{sec:technology}. Models are \emph{virtual models}. They are
  composed of \emph{concepts}, being themselves concepts. Concepts may have
  \emph{roles} and \emph{behaviors} (actions one can perform).

  In order to provide the graphical tools, our solution also uses \emph{model
  slots} to connect some concept instances to external elements (in our case:
  instances of diagram from our diagraming TA). The use of slots has been
  described in the section~\ref{sec:model} and is illustrated by bold circles in
  the Figure~\ref{fig:ToolingArchitecture}).

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation.

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models.  Note that in our approach, the
  metamodels are built in an ad hoc way together with the models
  (co-construction), therefore the user also validates on an ad hoc basis.

@@

  %\noteJC{pas trop clair pour moi pour le moment -> N/A ? }
  Due to the nature of our approach and its level-agnosticism, \emph{deep characterization} does not 
  apply to our solution. Such a mechanism could probably be encoded by specific behaviors, however, 
  it would not be a generic mechanism, making it difficult to reuse for another problem.

@@

  Due to the level-agnostic aspect of our approach, we could insert a new
  level, for example. It would consist in creating new concepts we would link
  to other ones. Then, we would add any necessary constraints on those concepts
  and on the relationships linking them to the other existing concepts. This type 
  of change can be done without much effort. On the other hand, making changes 
  to the existing models could be difficult. For example, given a requirement, 
  an instance cannot be changed into a concept easily. This case could occur if 
  the challenge had refined the specification by requiring specific design tasks. 
  In this context, we would have to transform \emph{Design Task} into a concept 
  whose instances could have been "using agile methodology".

@@

%\todo[inline]{recommended discussion aspects}
\subsection{Tools}

@@

As described in the section~\ref{subsec:tooling}, our solution is fully tooled with
Openflexo.  We use the Openflexo infrastructure to build models and metamodels
together, and to provide dedicated tooling to edit a Process type, to enact a
process and to execute it. Being able to provide a solution and its associated
tools quickly and easily is a noticeable feature.  We could also have taken
advantage of the model federation to connect our solution to other tools
dedicated to process edition and to process execution (\eg a \BPMN engine).
However we already had all the necessary components to provide the
aforementioned graphical tools.

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked. Thus, a part of
the verification relies on the fact that users add behaviors when they build
the metamodels and the models.
